---
title: "ShapleyForest on Real Data:  ALL Leukemia"
author: "Timothy Shen"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Shapley Forest on Real Data: ALL Leukemia}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 9, fig.height = 5)
```

# Introduction

This vignette demonstrates the full Shapley Forest (SF) workflow on the
**Acute Lymphoblastic Leukemia (ALL)** microarray dataset. The dataset is a canonical
high-dimensional genomics benchmark originally in `ALL` package.



The vignette proceeds in four stages:

1. Data loading and summary
2. WGCNA — soft-threshold selection, module detection, dendrogram
3. Single SF run and output exploration
4. Nested cross-validation comparing all six SF variants with joint
   hyperparameter tuning

---

# Data

The preprocessed dataset ships with the package — no Bioconductor installation
needed.

```{r data}
if (!require("devtools", quietly = TRUE))
    install.packages("devtools")
library(devtools)
devtools::install_github("timothyzshen/shapleyforest")

library(shapleyforest)

data(leukemia)

leukemia_X <- leukemia$X
leukemia_y <- leukemia$y

cat("Samples :", nrow(leukemia_X), "\n")
cat("Features:", ncol(leukemia_X), "\n")
cat("Outcome :", paste(names(table(leukemia_y)), collapse = " / "),
    "→", paste(table(leukemia_y), collapse = " / "), "\n")
```

---

# WGCNA Module Detection

Before running SF we perform WGCNA ourselves so we can inspect the network
structure.  `wsf()` can run WGCNA internally, but doing it manually lets us
examine — and justify — the soft-threshold power and the resulting modules.

```{r wgcna-load}
library(WGCNA)
options(stringsAsFactors = FALSE)
enableWGCNAThreads()
```

## Soft-threshold power selection

WGCNA constructs a weighted gene co-expression network by raising pairwise
correlations to a soft-thresholding power β.  The chosen power should make
the resulting network approximately **scale-free** (R² ≥ 0.85) while keeping
**mean connectivity** reasonably high (not near zero).

```{r soft-threshold, eval=false}
powers <- c(1:10, seq(12, 20, by = 2))

sft <- pickSoftThreshold(
  leukemia_X,
  powerVector  = powers,
  networkType  = "signed hybrid",
  verbose      = 0
)
```

```{r soft-threshold-plot, eval=false}
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Left: scale-free topology fit (R²) vs soft power. Right: mean network connectivity vs soft power. Choose the lowest power where R² ≥ 0.85 (dashed line)."

par(mfrow = c(1, 2), mar = c(5, 5, 3, 1))

# ── R² (scale-free fit) ────────────────────────────────────────────────────────
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft-threshold power (β)",
     ylab = expression(paste("Scale-free topology fit R"^2)),
     main = "Scale-free fit",
     type = "n", ylim = c(0, 1))
abline(h = 0.85, lty = 2, col = "tomato", lwd = 1.5)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, col = "#1F77B4", cex = 0.9, font = 2)

# ── Mean connectivity ──────────────────────────────────────────────────────────
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft-threshold power (β)",
     ylab = "Mean connectivity",
     main = "Mean connectivity",
     type = "n")
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers, col = "#D62728", cex = 0.9, font = 2)
```

From the plots we choose **β = 12** — the lowest power where R² ≥ 0.85 and
mean connectivity is still distinguishable from zero.

## Module detection and dendrogram

```{r wgcna-modules, eval=false}
SOFT_POWER <- 12L

net <- blockwiseModules(
  leukemia_X,
  power             = SOFT_POWER,
  networkType       = "signed hybrid",
  TOMType           = "signed",
  minModuleSize     = 10L,
  mergeCutHeight    = 0.25,
  numericLabels     = FALSE,
  pamRespectsDendro = FALSE,
  saveTOMs          = FALSE,
  verbose           = 0
)

table(net$colors)   # grey = unassigned (treated as independent by SF)
```

```{r wgcna-dendro, eval=false}
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "Gene dendrogram with module colour bar. Grey = unassigned features — these form the independent pool in SF's split shadow mode."

plotDendroAndColors(
  net$dendrograms[[1]],
  net$colors[net$blockGenes[[1]]],
  "Module",
  dendroLabels = FALSE,
  hang         = 0.03,
  addGuide     = TRUE,
  guideHang    = 0.05,
  main         = "Gene dendrogram and module colours"
)
```

### Module membership summary

```{r module-summary, eval=false}
mod_tbl <- as.data.frame(table(Module = net$colors))
mod_tbl$Independent <- mod_tbl$Module == "grey"
mod_tbl[order(mod_tbl$Freq, decreasing = TRUE), ]
```

Grey-labelled features are **unassigned by WGCNA** and treated as the
independent pool in SF's split shadow-stability mode.  A non-trivial
independent pool (> 5 % of features) validates the split architecture.

### Extract module membership for SF

```{r module-membership, eval=false}
# Convert WGCNA colour labels → named character vector expected by sf()
module_membership <- net$colors
names(module_membership) <- colnames(leukemia_X)

# Recode grey → "independent" so SF recognises the independent pool
module_membership[module_membership == "grey"] <- "independent"
```

---

# Single SF Run

We fit SF once on the full dataset using the module membership derived above
and the pure-R backend so no Python environment is required.

```{r single-run, eval=false}
set.seed(42)
fit <- sf(
  X                = leukemia_X,
  y                = leukemia_y,
  module_membership = module_membership,
  backend          = "R",
  r_shap           = "fastshap",
  verbose          = 1L
)
```

## Stable set

```{r stable-set, eval=false}
cat("Stable features selected:", length(fit$stable_features), "\n")
head(sort(fit$stability_freq, decreasing = TRUE), 20)
```

## Stability plot

```{r plot-stability, eval=false}
#| fig-cap: "Shadow-win frequency per feature. Filled points exceed the stability threshold and enter the stable set."
plot_stability(fit)
```

The dashed horizontal line is the tuned `pi_thr`.  The right panel shows the
independent pool with its own elbow-detected threshold.

## Feature importance

```{r plot-importance, eval=false}
#| fig-cap: "Mean |SHAP| for the top 20 stable features."
plot_importance(fit, top_n = 20)
```

---

# Nested Cross-Validation

## Design

We compare **six SF variants** formed by two shadow modes × three DML options:

| Variant | Shadow mode | DML |
|---------|------------|-----|
| `split` | split | none |
| `split_dml` | split | standard |
| `split_mdml` | split | marginal |
| `within` | within_module | none |
| `within_dml` | within_module | standard |
| `within_mdml` | within_module | marginal |

**Outer loop (k = 5):** five held-out test folds that are never seen during
tuning.  Provides the unbiased AUC estimates used for variant comparison.

**Inner loop (k = 3):** three-fold CV within each outer training fold.
Used to jointly tune three hyperparameters:

| Parameter | Grid | Default |
|-----------|------|---------|
| `screen_control(drop_fraction)` | 0.15, 0.25, 0.35 | 0.25 |
| `select_control(drop_fraction)` | 0.05, 0.10, 0.20 | 0.10 |
| `select_control(pi_thr)` | 0.40, 0.50, 0.60, 0.70 | 0.60 |

36 combinations per inner fold.  Selection criterion: mean inner AUC.
All other parameters remain at their package defaults.

## Setup

```{r cv-setup, eval=false}
library(pROC)
library(caret)
library(future.apply)
plan(multisession)

VARIANTS <- list(
  split       = list(shadow_mode = "split",         dml = FALSE),
  split_dml   = list(shadow_mode = "split",         dml = TRUE),
  split_mdml  = list(shadow_mode = "split",         dml = "marginal"),
  within      = list(shadow_mode = "within_module", dml = FALSE),
  within_dml  = list(shadow_mode = "within_module", dml = TRUE),
  within_mdml = list(shadow_mode = "within_module", dml = "marginal")
)

GRID <- expand.grid(
  screen_drop = c(0.15, 0.25, 0.35),
  select_drop = c(0.05, 0.10, 0.20),
  pi_thr      = c(0.40, 0.50, 0.60, 0.70),
  stringsAsFactors = FALSE
)

set.seed(42)
outer_folds <- caret::createFolds(leukemia_y, k = 5, returnTrain = FALSE)
```

## Fit-and-evaluate helper

```{r cv-helper, eval=false}
fit_eval <- function(X_tr, y_tr, X_te, y_te, variant, grid_row) {

  sc  <- screen_control(drop_fraction = grid_row$screen_drop)
  sl  <- select_control(
    drop_fraction = grid_row$select_drop,
    pi_thr        = grid_row$pi_thr,
    shadow_mode   = variant$shadow_mode
  )

  # WGCNA on this fold's training data
  net_fold <- blockwiseModules(
    X_tr, power = SOFT_POWER, networkType = "signed hybrid",
    minModuleSize = 10L, mergeCutHeight = 0.25,
    numericLabels = FALSE, saveTOMs = FALSE, verbose = 0
  )
  mm_fold <- net_fold$colors
  mm_fold[mm_fold == "grey"] <- "independent"

  fit <- tryCatch(
    sf(X_tr, y_tr,
       module_membership = mm_fold,
       screen_params     = sc,
       select_params     = sl,
       dml               = variant$dml,
       backend           = "R", r_shap = "permutation",
       verbose           = 0L),
    error = function(e) NULL
  )

  stable <- if (!is.null(fit)) fit$stable_features else character(0)
  if (length(stable) == 0L)
    return(list(auc = NA_real_, brier = NA_real_,
                n_stable = 0L, stable = character(0)))

  rf   <- ranger::ranger(
    y ~ ., probability = TRUE, num.trees = 300L, seed = 1L,
    data = data.frame(y = factor(y_tr), X_tr[, stable, drop = FALSE])
  )
  prob <- predict(rf,
    data.frame(X_te[, stable, drop = FALSE]))$predictions[, 2]

  list(
    auc      = as.numeric(pROC::auc(y_te, prob, quiet = TRUE)),
    brier    = mean((prob - y_te)^2),
    n_stable = length(stable),
    stable   = stable
  )
}
```

## Nested CV loop

```{r nested-cv, eval=false}
nested_cv_variant <- function(variant_name) {
  variant <- VARIANTS[[variant_name]]

  outer_results <- lapply(seq_along(outer_folds), function(o) {

    test_idx  <- outer_folds[[o]]
    train_idx <- setdiff(seq_along(leukemia_y), test_idx)
    X_tr <- leukemia_X[train_idx, ];  y_tr <- leukemia_y[train_idx]
    X_te <- leukemia_X[test_idx,  ];  y_te <- leukemia_y[test_idx]

    # ── Inner loop: tune all three params jointly ──────────────────────────
    inner_folds <- caret::createFolds(y_tr, k = 3L, returnTrain = FALSE)

    inner_auc <- future_sapply(seq_len(nrow(GRID)), function(g) {
      mean(vapply(inner_folds, function(iv) {
        i_te <- iv; i_tr <- setdiff(seq_along(y_tr), iv)
        fit_eval(X_tr[i_tr, ], y_tr[i_tr],
                 X_tr[i_te, ], y_tr[i_te],
                 variant, GRID[g, ])$auc
      }, numeric(1L)), na.rm = TRUE)
    }, future.seed = TRUE)

    best_row <- GRID[which.max(inner_auc), ]

    # ── Outer: refit on full outer-train ──────────────────────────────────
    res <- fit_eval(X_tr, y_tr, X_te, y_te, variant, best_row)

    list(
      auc              = res$auc,
      brier            = res$brier,
      n_stable         = res$n_stable,
      stable           = res$stable,
      best_screen_drop = best_row$screen_drop,
      best_select_drop = best_row$select_drop,
      best_pi_thr      = best_row$pi_thr
    )
  })

  # ── Jaccard across outer folds ───────────────────────────────────────────
  sets     <- lapply(outer_results, `[[`, "stable")
  jac_fn   <- function(a, b) length(intersect(a,b)) / max(length(union(a,b)), 1L)
  jac_mean <- mean(combn(length(sets), 2,
                         function(i) jac_fn(sets[[i[1]]], sets[[i[2]]]),
                         simplify = TRUE))

  fold_df <- data.frame(
    fold             = seq_along(outer_results),
    auc              = vapply(outer_results, `[[`, numeric(1), "auc"),
    brier            = vapply(outer_results, `[[`, numeric(1), "brier"),
    n_stable         = vapply(outer_results, `[[`, numeric(1), "n_stable"),
    best_pi_thr      = vapply(outer_results, `[[`, numeric(1), "best_pi_thr"),
    best_screen_drop = vapply(outer_results, `[[`, numeric(1), "best_screen_drop"),
    best_select_drop = vapply(outer_results, `[[`, numeric(1), "best_select_drop")
  )

  list(
    variant  = variant_name,
    fold_df  = fold_df,
    auc_mean = mean(fold_df$auc, na.rm = TRUE),
    auc_sd   = sd(  fold_df$auc, na.rm = TRUE),
    brier    = mean(fold_df$brier, na.rm = TRUE),
    n_stable = mean(fold_df$n_stable, na.rm = TRUE),
    jaccard  = jac_mean,
    sets     = sets
  )
}

all_cv <- future_lapply(names(VARIANTS), nested_cv_variant,
                        future.seed = TRUE)
names(all_cv) <- names(VARIANTS)
```

## Results

### Summary table

```{r cv-table, eval=false}
summary_tbl <- do.call(rbind, lapply(all_cv, function(r) {
  data.frame(
    Variant         = r$variant,
    `AUC mean±SD`   = sprintf("%.3f ± %.3f", r$auc_mean, r$auc_sd),
    Brier           = round(r$brier,   3),
    `Mean |U|`      = round(r$n_stable, 1),
    Jaccard         = round(r$jaccard,  3),
    `Best pi_thr`   = names(sort(table(r$fold_df$best_pi_thr),
                                 decreasing = TRUE))[1],
    `Best sc_drop`  = names(sort(table(r$fold_df$best_screen_drop),
                                 decreasing = TRUE))[1],
    `Best sl_drop`  = names(sort(table(r$fold_df$best_select_drop),
                                 decreasing = TRUE))[1],
    check.names = FALSE
  )
}))
knitr::kable(summary_tbl, align = "lcccccc",
             caption = "Nested CV summary across five outer folds.")
```

### Pairwise variant comparison

Five outer folds give five paired AUC values per variant.  We test all 15
pairs with a paired Wilcoxon test and apply Benjamini-Hochberg correction.

```{r pairwise, eval=false}
pairs <- combn(names(VARIANTS), 2, simplify = FALSE)

pw <- do.call(rbind, lapply(pairs, function(p) {
  a <- all_cv[[p[1]]]$fold_df$auc
  b <- all_cv[[p[2]]]$fold_df$auc
  wt <- wilcox.test(a, b, paired = TRUE, exact = FALSE)
  data.frame(`Method A` = p[1], `Method B` = p[2],
             `ΔAUC`     = round(mean(a - b, na.rm = TRUE), 3),
             `p-value`  = round(wt$p.value, 4),
             check.names = FALSE)
}))
pw$`p-adj (BH)` <- round(p.adjust(pw$`p-value`, method = "BH"), 4)
knitr::kable(pw, caption = "Pairwise Wilcoxon tests on outer-fold AUC.")
```

### Jaccard stability

```{r jaccard-plot, eval=false}
#| fig-cap: "Mean Jaccard similarity of stable sets across the five outer folds. High values indicate SF selects the same genes regardless of which patients it trains on."
library(ggplot2)

jac_df <- data.frame(
  Variant = names(all_cv),
  Jaccard = vapply(all_cv, `[[`, numeric(1), "jaccard")
)

ggplot(jac_df, aes(x = reorder(Variant, Jaccard), y = Jaccard)) +
  geom_col(fill = "#1F77B4", width = 0.6) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  coord_flip() +
  labs(title    = "Stable-set consistency across outer folds",
       subtitle = "Dashed line = 0.5 reference",
       x = NULL, y = "Mean pairwise Jaccard") +
  theme_bw()
```

### Biological validation

```{r bio, eval=false}
# Best-performing variant by mean outer AUC
best_v  <- names(which.max(vapply(all_cv, `[[`, numeric(1), "auc_mean")))
cat("Best variant:", best_v, "\n")

# Union of stable sets across the five outer folds
stable_union <- Reduce(union, all_cv[[best_v]]$sets)
cat("Union stable set size:", length(stable_union), "\n")

# A few well-known B-cell / T-cell markers from the HGU95Av2 platform
# (probe-set IDs; will be present only if they fall in the top-500 by variance)
known_bcell <- c("1303_at",  "32069_at")   # CD19, PAX5 proxies
known_tcell <- c("38319_at", "1827_s_at")  # CD3D, CD3E proxies
cat("B-cell markers recovered:", sum(known_bcell %in% stable_union),
    "/", length(known_bcell), "\n")
cat("T-cell markers recovered:", sum(known_tcell %in% stable_union),
    "/", length(known_tcell), "\n")
```

---

# Reporting Checklist

| Item | Source |
|------|--------|
| AUC mean ± SD per variant | `summary_tbl` |
| Brier score | `summary_tbl` |
| Mean stable set size |U| | `summary_tbl` |
| Most-chosen tuned parameters | mode of `fold_df$best_*` columns |
| Pairwise variant comparison | paired Wilcoxon + BH table |
| Jaccard across outer folds | bar chart |
| Soft-threshold power justification | R² + connectivity plots |
| Biological overlap | union stable set vs known markers |

---

# Session Info

```{r session}
sessionInfo()
```
