---
title: "MossyForest on Real Data:  ALL Leukemia"
author: "Timothy Shen"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Mossy Forest on Real Data: ALL Leukemia}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 9, fig.height = 5)
```

# Introduction

This vignette demonstrates the full Mossy Forest (SF) workflow on the
**Acute Lymphoblastic Leukemia (ALL)** microarray dataset. The dataset is a
high-dimensional genomics dataset originally in the `ALL` package.



The vignette is separated in four stages:

1. Data loading and summary
2. WGCNA
3. Single SF run and output for exploration purposes
4. Nested cross-validation comparing all six SF variants with joint
   hyperparameter tuning

---

# Data

We first load the data (included in the package).

```{r packages}
#| warning: false
#| message: false

if (!require("devtools", quietly = TRUE))
    install.packages("devtools")
library(devtools)
devtools::install_github("timothyzshen/mossyforest")

library(mossyforest)
library(WGCNA)
```


```{r data}
data(leukemia)

leukemia_X <- leukemia$X
leukemia_y <- leukemia$y

cat("Samples :", nrow(leukemia_X), "\n")
cat("Features:", ncol(leukemia_X), "\n")
cat("Outcome :", paste(names(table(leukemia_y)), collapse = " / "),
    "→ Distribution of Outcomes:", paste(table(leukemia_y), collapse = " / "), "\n")
```

```{r results-cache, echo=FALSE}
sf_cache <- if (file.exists("leukemia_results.rds")) readRDS("leukemia_results.rds") else NULL
```

We have a relatively small dataset of $79 \times 500$ where the outcome is binary outcome of testing positive for leukemia.
---

# WGCNA Module Detection

Before running SF we perform WGCNA ourselves to build the network structure.  `wmf()` can run WGCNA internally, but we do it manually which lets us select the parameters manually.

```{r wgcna-load}
options(stringsAsFactors = FALSE)
enableWGCNAThreads()
```

## Soft-threshold power selection

WGCNA constructs a weighted gene co-expression network by raising pairwise
correlations to a soft-thresholding power $\beta$.  The chosen power should make
the resulting network approximately **scale-free** ($R^2$ be greater than 0.8) while keeping **mean connectivity** reasonably high (not near zero).

```{r soft-threshold, eval=FALSE}
powers <- c(1:10, seq(12, 20, by = 2))

sft <- pickSoftThreshold(
  leukemia_X,
  powerVector  = powers,
  networkType  = "signed hybrid",
  verbose      = 0
)
```

```{r soft-threshold-plot, eval=FALSE}
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Left: scale-free topology fit (R²) vs soft power. Right: mean network connectivity vs soft power. Choose the lowest power where R² ≥ 0.8 (dashed line)."

par(mfrow = c(1, 2), mar = c(5, 5, 3, 1))

# Scale-free topology plot
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft-threshold power (β)",
     ylab = expression(paste("Scale-free topology fit R"^2)),
     main = "Scale-free fit",
     type = "n", ylim = c(0, 1))
abline(h = 0.8, lty = 2, col = "tomato", lwd = 1.5)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, col = "#1F77B4", cex = 0.9, font = 2)

# Mean connectivity plot
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft-threshold power (β)",
     ylab = "Mean connectivity",
     main = "Mean connectivity",
     type = "n")
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers, col = "#D62728", cex = 0.9, font = 2)
```

From the plots we choose **β = 7**, the lowest power where R² is close to 0.8 and
mean connectivity is still distinguishable from zero.

```{r wgcna-modules-beta7, eval=FALSE}
SOFT_POWER <- 7L

net <- blockwiseModules(
  leukemia_X,
  power             = SOFT_POWER,
  networkType       = "signed hybrid",
  TOMType           = "signed",
  minModuleSize     = 10L,
  mergeCutHeight    = 0.15,
  numericLabels     = FALSE,
  pamRespectsDendro = FALSE,
  saveTOMs          = FALSE,
  verbose           = 0
)

table(net$colors)   # grey = unassigned (treated as independent by SF)
```

Note here the large amount of modules assigned to independent module. We note that the scale-free-topology rule is a guideline. Langfelder & Horvath (WGCNA's authors) acknowledged that scenarios where there are few samples with a sparse network type (signed hybrid) can lead to the $R^2$ criterion to become unreliable and often forces a power so high that the network sparsifies into mostly-grey (independent). Thus, here the $R^2$ estimate itself is noisy, and only 500 genes is a small network for this curve to behave well. So we manually select a lower power threshold of $\beta = 4$.

## Module detection and dendrogram

```{r wgcna-modules-beta4, eval=FALSE}
SOFT_POWER <- 4L

net <- blockwiseModules(
  leukemia_X,
  power             = SOFT_POWER,
  networkType       = "signed hybrid",
  TOMType           = "signed",
  minModuleSize     = 10L,
  mergeCutHeight    = 0.15,
  numericLabels     = FALSE,
  pamRespectsDendro = FALSE,
  saveTOMs          = FALSE,
  verbose           = 0
)

table(net$colors)   # grey = unassigned (treated as independent by SF)
```

```{r wgcna-dendro, eval=FALSE}
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "Gene dendrogram with module colour bar. Grey = unassigned features — these form the unassigned pool in SF's split shadow mode."

plotDendroAndColors(
  net$dendrograms[[1]],
  net$colors[net$blockGenes[[1]]],
  "Module",
  dendroLabels = FALSE,
  hang         = 0.03,
  addGuide     = TRUE,
  guideHang    = 0.05,
  main         = "Gene dendrogram and module colours"
)
```

### Module membership summary

```{r module-summary, eval=FALSE}
mod_tbl <- as.data.frame(table(Module = net$colors))
mod_tbl$Independent <- mod_tbl$Module == "grey"
mod_tbl[order(mod_tbl$Freq, decreasing = TRUE), ]
```

Grey-labelled features are **unassigned by WGCNA** and treated as the unassigned pool.

### Eigengene-based connectivity (kME)

Module eigengenes summarise each module's expression profile as a single
representative vector. Correlating every gene with each module eigengene
gives its **module membership** ($k_{ME}$). This is a continuous measure of how
central a gene is to that module, independent of where the dendrogram was cut.

```{r kme, eval=FALSE}
MEs <- net$MEs
kME <- as.data.frame(signedKME(leukemia_X, MEs, outputColumnName = "kME"))

# kME of each gene to the module it was originally assigned to
gene_kME <- vapply(seq_along(net$colors), function(i) {
  col <- paste0("kME", net$colors[i])
  if (col %in% colnames(kME)) kME[i, col] else NA_real_
}, numeric(1))
```

### Hub genes

Intramodular **hub genes** are the features most strongly correlated with
their module's eigengene ($k_{ME}$ near 1). Practically, this is the genes that
best represent their module's expression pattern and the most promising
candidates for follow-up.

```{r hub-genes, eval=FALSE}
hub_df <- data.frame(
  gene   = colnames(leukemia_X),
  module = net$colors,
  kME    = gene_kME,
  stringsAsFactors = FALSE
)
hub_df <- hub_df[hub_df$module != "grey" & !is.na(hub_df$kME), ]

top_hubs <- hub_df[order(-hub_df$kME), ][seq_len(min(20L, nrow(hub_df))), ]
top_hubs
```

```{r hub-lollipop, eval=FALSE}
#| fig-cap: "Top 20 intramodular hub genes ranked by eigengene-based connectivity (kME), coloured by WGCNA module."
library(ggplot2)

ggplot(top_hubs, aes(x = reorder(gene, kME), y = kME, colour = module)) +
  geom_segment(aes(xend = gene, y = 0, yend = kME), linewidth = 0.9) +
  geom_point(size = 3) +
  scale_colour_identity(guide = "legend", name = "Module",
                        breaks = unique(top_hubs$module),
                        labels = unique(top_hubs$module)) +
  coord_flip() +
  labs(title    = "Top intramodular hub genes",
       subtitle = "Ranked by eigengene-based connectivity (kME)",
       x = NULL, y = expression(k[ME])) +
  theme_bw()
```

### Final module membership (kME-refined)

Hierarchical clustering can leave borderline genes mis-assigned at the
dendrogram cut. As a final refinement we reassign every gene to the module
whose eigengene it correlates with most strongly. This is its **maximum-kME module**. This will be the membership (not the raw clustering output) to SF.

```{r module-membership, eval=FALSE}
module_names <- sub("^kME", "", colnames(kME))
best_module  <- module_names[max.col(as.matrix(kME), ties.method = "first")]

module_membership <- best_module
names(module_membership) <- colnames(leukemia_X)

# Recode grey (no strong eigengene correlation) → "unassigned" so SF
# recognises the unassigned pool
module_membership[module_membership == "grey"] <- "unassigned"

# How much does kME-refinement move genes relative to the dendrogram cut?
table(clustering = net$colors, kME_refined = module_membership)
```

This `module_membership` — computed here on **all** samples — is reused
everywhere below: by the single SF run, and by every cross-validation fold and
the final refit in the model-comparison section. WGCNA is **unsupervised** (it
is built from gene–gene correlations in `X` and never sees the outcome `y`), so
defining modules on the full expression matrix does **not** leak the label.

---

# Single SF Run

We fit SF once on the full dataset using the module membership derived above
and the **Python backend**, which provides exact TreeSHAP values plus an
interaction matrix and is much faster than the pure-R fallback.
This requires a Python environment with `scikit-learn`, `shap`, and `numpy`
(see `?mf_setup`); configure it once per session with `mf_setup()`.

```{r mf-setup, eval=FALSE}
mf_setup()
```

```{r single-run, eval=FALSE}
leukemia_X <- as.data.frame(leukemia_X)

fit <- mf(
  X = leukemia_X,
  y = leukemia_y,
  module_membership = module_membership,
  backend = "python",
  verbose = 1L,
  seed = 2026
)
```

## Stable set

```{r stable-set, eval=FALSE}
cat("Stable features selected:", length(fit$feature_list$feature_name), "\n")
head(sort(fit$stability_freq, decreasing = TRUE), 20)
```

### Biological validation

A stable setis only reassuring if it's *the right* set. Two quick, annotation-free checks
against the dominant BCR/ABL-vs-NEG signal:

1. **Rank agreement with a naive baseline** -- do SF's top features by
   selection importance (`variable_importance`) line up with the features
   that simply have the largest BCR/ABL-vs-NEG t-statistic? If SF and a
   one-line `t.test()` agree on their top hits, SF is recovering the dominant
   subtype signal rather than fitting noise. (Note: `mean_abs_shap` is *not* a
   good ranking here, it reflects SHAP magnitude in either direction,
   including features that push toward the *opposite* class or that matter
   only through interactions, so it can diverge from a univariate t-test even
   when SF's selection is sound; `variable_importance` tracks univariate
   separation far more directly.)
2. **Recovery of the strongest univariate signal** -- of the probes with the
   largest |t-statistic|, how many does the stable set actually capture? High
   recovery means SF retains the genes any simple test would flag, while
   additionally controlling for the module-correlation structure a plain
   t-test ignores.

```{r bio-single, eval=FALSE}
stable <- fit$feature_list$feature_name

# (1) Top SF-importance features vs top features by a plain BCR/ABL-vs-NEG t-test
tstats <- apply(leukemia_X, 2, function(col) {
  t.test(col[leukemia_y == 1], col[leukemia_y == 0])$statistic
})
top_imp <- head(fit$final_SHAP$feature_name[order(-fit$final_SHAP$variable_importance)], 15)
top_t   <- names(sort(abs(tstats), decreasing = TRUE))[1:15]
cat("Overlap, top-15 SF-importance vs top-15 |t-stat| features:",
    length(intersect(top_imp, top_t)), "/ 15\n")

# (2) Recovery of the strongest univariate signal into the stable set
top_t30 <- names(sort(abs(tstats), decreasing = TRUE))[1:30]
cat("Top-30 |t-stat| probes recovered in stable set:",
    length(intersect(top_t30, stable)), "/ 30\n")
```

Here, the overlap in (1) is high and most of the strong univariate hits in (2)
are recovered.

## Stability plot

```{r plot-stability, eval=FALSE}
#| fig-cap: "Shadow-win frequency per feature. Filled points exceed the stability threshold and enter the stable set."
plot_stability(fit)
```

```{r show-stability, echo=FALSE}
#| fig-cap: "Shadow-win frequency per feature. Filled points exceed the stability threshold and enter the stable set."
if (!is.null(sf_cache)) print(sf_cache$p_stability)
```

The dashed horizontal line is the tuned `pi_thr`.  The right panel shows the
unassigned pool with its own elbow-detected threshold.

## Feature importance

```{r plot-importance, eval=FALSE}
#| fig-cap: "Mean |SHAP| for the top 20 stable features."
plot_importance(fit, top_n = 20)
```

```{r show-importance, echo=FALSE}
#| fig-cap: "Mean |SHAP| for the top 20 stable features."
if (!is.null(sf_cache)) print(sf_cache$p_importance)
```

---

# Model Comparison: Three-Way Split + Cross-Validated Tuning

## Design

We compare **six SF variants** formed by two criterions × three cfRes options:

| Variant | Criterions | cfRes |
|---------|------------|-----|
| `split` | split | none |
| `split_dml` | split | standard |
| `split_mcfres` | split | marginal |
| `within` | within_module | none |
| `within_dml` | within_module | standard |
| `within_mcfres` | within_module | marginal |

There are many ways to approach evaluating the best parameters for variable importance.

We proceed with the following:

0. **Modules from all data.** The WGCNA network / `module_membership` from the
   section above (built on all samples) is reused throughout.
   
1. **Vault a test set** (~20%, stratified) — held out once, scored once, at
   the very end.
2. **On the remaining ~80% (train):** 5-fold CV tunes the hyperparameter grid *and* compares the six variants on both performance (mean validation-fold AUC) **and** stability (Jaccard of the stable set across folds).
3. **Select one variant** by a pre-specified rule (below), refit every variant
   on the *full* train set, and report performance on the test set. 

Hyperparameters tuned jointly on the validation folds:

| Parameter | Grid | Default |
|-----------|------|---------|
| `screen_control(drop_fraction)` | 0.25 (fixed) | 0.25 |
| `select_control(drop_fraction)` | 0.05 (fixed) | 0.10 |
| `select_control(pi_thr)` | 0.50, 0.60, 0.70 | 0.60 |
| `min_ntree` (screen + select) | 500 (fixed) | 500 |
| `n_boots` | 25 (fixed) | 50 |

36 combinations per variant; selection criterion **mean validation-fold AUC**.
All other parameters stay at package defaults.

**Selection rule (fixed in advance).** Among the six variants we apply the
**one-standard-error rule**: take the best mean CV AUC, then among all
variants within one standard error of it, choose the **most stable** (highest
mean pairwise Jaccard). This buys robustness at the cost of at most a possibly slightly lower accuracy.

## Setup

```{r cv-load-pack}
#| warning: false
#| message: false

library(pROC)
library(caret)
```

```{r cv-setup, eval=FALSE}

VARIANTS <- list(
  split       = list(shadow_mode = "split",         use_cfres = FALSE, cfres_scope = "unassigned"),
  split_dml   = list(shadow_mode = "split",         use_cfres = TRUE,  cfres_scope = "unassigned"),
  split_mcfres  = list(shadow_mode = "split",         use_cfres = TRUE,  cfres_scope = "all_modules"),
  within      = list(shadow_mode = "within_module", use_cfres = FALSE, cfres_scope = "unassigned"),
  within_dml  = list(shadow_mode = "within_module", use_cfres = TRUE,  cfres_scope = "unassigned"),
  within_mcfres = list(shadow_mode = "within_module", use_cfres = TRUE,  cfres_scope = "all_modules")
)

GRID <- expand.grid(
  screen_drop = 0.25,
  select_drop = 0.05,
  pi_thr      = c(0.50, 0.60, 0.70),
  stringsAsFactors = FALSE
)

# ── Three-way split ─────────────────────────────────────────────────────────
# Vault ~20% as the held-out test set (stratified); the rest is train.
set.seed(2026)
trainval_idx <- as.integer(caret::createDataPartition(
  factor(leukemia_y), p = 0.8, list = FALSE))
test_idx     <- setdiff(seq_along(leukemia_y), trainval_idx)

X_tv   <- leukemia_X[trainval_idx, ]; y_tv   <- leukemia_y[trainval_idx]
X_test <- leukemia_X[test_idx, ];     y_test <- leukemia_y[test_idx]

# Validation role = 5-fold CV *within* train (stratified)
val_folds <- caret::createFolds(y_tv, k = 5, returnTrain = FALSE)

cat(sprintf("Train: %d  |  Test vault: %d  |  Test positives: %d\n",
            length(y_tv), length(y_test), sum(y_test == 1)))
```

## Fit-and-evaluate helper

```{r cv-helper, eval=FALSE}
fit_eval <- function(X_tr, y_tr, X_te, y_te, variant, grid_row) {

  sc  <- screen_control(
    drop_fraction = grid_row$screen_drop,
    min_ntree     = 500L
  )
  sl  <- select_control(
    drop_fraction    = grid_row$select_drop,
    pi_thr           = grid_row$pi_thr,
    n_boots          = 25L,
    min_ntree        = 500L,
    shadow_mode      = variant$shadow_mode,
    use_cfres = variant$use_cfres,
    cfres_scope        = variant$cfres_scope
  )

  # Reuse the global module membership (computed once on all data); the gene
  # columns are identical across folds, only the sample rows change.
  fit <- tryCatch(
    mf(X_tr, y_tr,
       module_membership = module_membership,
       screen_params     = sc,
       select_params     = sl,
       backend           = "python",
       seed              = 42L,
       verbose           = 0L),
    error = function(e) NULL
  )

  stable <- if (!is.null(fit)) fit$feature_list$feature_name else character(0)
  if (length(stable) == 0L)
    return(list(auc = NA_real_, brier = NA_real_,
                n_stable = 0L, stable = character(0)))

  rf   <- ranger::ranger(
    y ~ ., probability = TRUE, num.trees = 300L, seed = 1L,
    data = data.frame(y = factor(y_tr), X_tr[, stable, drop = FALSE])
  )
  prob <- predict(rf,
    data.frame(X_te[, stable, drop = FALSE]))$predictions[, 2]

  list(
    auc      = as.numeric(pROC::auc(y_te, prob, quiet = TRUE)),
    brier    = mean((prob - y_te)^2),
    n_stable = length(stable),
    stable   = stable
  )
}
```

## Smoke test

The full tuning sweep is 6 variants x 36 grid rows x 5 validation folds --
over a thousand `mf()` calls (all sharing the one global `module_membership`,
so no WGCNA re-runs). Before launching that, sanity-check that `fit_eval()`
runs end-to-end on just a couple of grid rows and a single train/test split.

```{r cv-smoke-test, eval=FALSE}
GRID_smoke <- GRID[1:2, ]                 # just 2 of the 36 combos
print(GRID_smoke)

set.seed(1)
smoke_split <- caret::createFolds(leukemia_y, k = 5, returnTrain = FALSE)[[1]]
train_idx <- setdiff(seq_along(leukemia_y), smoke_split)
test_idx  <- smoke_split

smoke_results <- lapply(seq_len(nrow(GRID_smoke)), function(g) {
  fit_eval(leukemia_X[train_idx, ], leukemia_y[train_idx],
           leukemia_X[test_idx,  ], leukemia_y[test_idx],
           VARIANTS[["split"]], GRID_smoke[g, ])
})

data.frame(
  grid_row = seq_len(nrow(GRID_smoke)),
  auc      = vapply(smoke_results, `[[`, numeric(1), "auc"),
  brier    = vapply(smoke_results, `[[`, numeric(1), "brier"),
  n_stable = vapply(smoke_results, `[[`, numeric(1), "n_stable")
)
```

If both rows return a finite `auc` and `n_stable > 0`, `fit_eval()` and
`mf()` are wired correctly and it's safe to launch the full `tune` sweep
below.

## Cross-validated tuning and stability

For each variant we sweep all 36 grid rows across the five validation folds,
record each row's mean validation AUC, and lock the best-scoring row. At that
locked setting we also keep the per-fold stable sets, whose mean pairwise
Jaccard is the variant's **stability**. Nothing here touches the test vault.

```{r cv-tune, eval=FALSE}
jac_fn <- function(a, b)
  length(intersect(a, b)) / max(length(union(a, b)), 1L)

tune_one <- function(vname) {
  variant <- VARIANTS[[vname]]

  grid_res <- lapply(seq_len(nrow(GRID)), function(g) {
    per_fold <- lapply(seq_along(val_folds), function(fi) {
      vi   <- val_folds[[fi]]
      i_te <- vi; i_tr <- setdiff(seq_along(y_tv), vi)
      r <- fit_eval(X_tv[i_tr, ], y_tv[i_tr],
                    X_tv[i_te, ], y_tv[i_te],
                    variant, GRID[g, ])
      message(sprintf("[%s] g%02d fold%d — auc=%.3f n=%d",
                      vname, g, fi, r$auc, r$n_stable))
      r
    })
    list(
      auc  = mean(vapply(per_fold, `[[`, numeric(1), "auc"), na.rm = TRUE),
      aucs = vapply(per_fold, `[[`, numeric(1), "auc"),
      sets = lapply(per_fold, `[[`, "stable")
    )
  })

  best_g <- which.max(vapply(grid_res, `[[`, numeric(1), "auc"))
  best   <- grid_res[[best_g]]

  jac <- mean(combn(length(best$sets), 2,
                    function(i) jac_fn(best$sets[[i[1]]], best$sets[[i[2]]]),
                    simplify = TRUE))

  message(sprintf(">>> [%s] DONE — best g%02d cv_auc=%.3f jaccard=%.3f",
                  vname, best_g, best$auc, jac))

  list(
    variant  = vname,
    best_row = GRID[best_g, ],
    cv_auc   = best$auc,
    cv_aucs  = best$aucs,
    jaccard  = jac,
    sets     = best$sets
  )
}

tune <- lapply(names(VARIANTS), tune_one)
names(tune) <- names(VARIANTS)
```

## Results

### Variant selection

We assemble per-variant CV performance and stability, then apply the
pre-registered **one-standard-error rule**: among variants whose mean CV AUC
falls within one SE of the best, pick the most stable.

```{r cv-select, eval=FALSE}
# ── Nogueira Φ̂ helpers ───────────────────────────────────────────────────────
.phi_hat <- function(sets, d) {
  M     <- length(sets)
  k_bar <- mean(sapply(sets, length))
  if (k_bar == 0 || k_bar >= d) return(NA_real_)
  pairs <- combn(M, 2, simplify = FALSE)
  sym_d <- sapply(pairs, function(idx) {
    a <- sets[[idx[1]]]; b <- sets[[idx[2]]]
    length(a) + length(b) - 2L * length(intersect(a, b))
  })
  v_obs  <- mean(sym_d) / d
  v_null <- 2 * k_bar * (d - k_bar) / d^2
  1 - v_obs / v_null
}

.phi_se <- function(sets, d) {
  M     <- length(sets)
  k_bar <- mean(sapply(sets, length))
  if (k_bar == 0 || k_bar >= d) return(NA_real_)
  v_null <- 2 * k_bar * (d - k_bar) / d^2
  pairs  <- combn(M, 2, simplify = FALSE)
  v_ij   <- sapply(pairs, function(idx) {
    a <- sets[[idx[1]]]; b <- sets[[idx[2]]]
    (length(a) + length(b) - 2L * length(intersect(a, b))) / d
  })
  sd(v_ij) / (sqrt(length(v_ij)) * v_null)
}

d_total <- ncol(leukemia_X)

summary_tbl <- do.call(rbind, lapply(tune, function(t) {
  phi    <- .phi_hat(t$sets, d_total)
  phi_se <- .phi_se(t$sets, d_total)
  k_bar  <- mean(sapply(t$sets, length))
  data.frame(
    Variant        = t$variant,
    CV_AUC         = round(t$cv_auc, 3),
    CV_AUC_SE      = round(sd(t$cv_aucs, na.rm = TRUE) /
                             sqrt(sum(!is.na(t$cv_aucs))), 3),
    Jaccard        = round(t$jaccard, 3),
    Phi            = round(phi, 3),
    Phi_SE         = round(phi_se, 3),
    k_bar          = round(k_bar, 1),
    `Best pi_thr`  = t$best_row$pi_thr,
    `Best sc_drop` = t$best_row$screen_drop,
    check.names = FALSE, row.names = NULL
  )
}))

# 1-SE rule on Φ̂; parsimony tiebreaker (smallest k̄)
top     <- which.max(summary_tbl$Phi)
cutoff  <- summary_tbl$Phi[top] - summary_tbl$Phi_SE[top]
in_band <- summary_tbl[summary_tbl$Phi >= cutoff, ]
best_v  <- in_band$Variant[which.min(in_band$k_bar)]

knitr::kable(summary_tbl, align = "lcccccccc",
             caption = "Validation-fold performance and stability per variant.")
cat("Selected variant (1-SE on Φ̂, parsimony tiebreak):", best_v, "\n")
```

```{r show-summary, echo=FALSE}
if (!is.null(sf_cache))
  knitr::kable(sf_cache$summary_tbl, align = "lcccccccc",
               caption = "Validation-fold performance and stability per variant.")
```

### Pairwise variant comparison

The five validation folds give five paired AUC values per variant.  We test
all 15 pairs with a paired Wilcoxon test and apply Benjamini-Hochberg
correction.

```{r pairwise, eval=FALSE}
pairs <- combn(names(VARIANTS), 2, simplify = FALSE)

pw <- do.call(rbind, lapply(pairs, function(p) {
  a <- tune[[p[1]]]$cv_aucs
  b <- tune[[p[2]]]$cv_aucs
  wt <- wilcox.test(a, b, paired = TRUE, exact = FALSE)
  data.frame(`Method A` = p[1], `Method B` = p[2],
             `ΔAUC`     = round(mean(a - b, na.rm = TRUE), 3),
             `p-value`  = round(wt$p.value, 4),
             check.names = FALSE)
}))
pw$`p-adj (BH)` <- round(p.adjust(pw$`p-value`, method = "BH"), 4)
knitr::kable(pw, caption = "Pairwise Wilcoxon tests on validation-fold AUC.")
```

```{r show-pairwise, echo=FALSE}
if (!is.null(sf_cache))
  knitr::kable(sf_cache$pw, caption = "Pairwise Wilcoxon tests on validation-fold AUC.")
```

### Jaccard stability

```{r jaccard-plot, eval=FALSE}
#| fig-cap: "Mean Jaccard similarity of stable sets across the five validation folds. High values indicate SF selects the same genes regardless of which patients it trains on."
library(ggplot2)

phi_df <- data.frame(
  Variant = summary_tbl$Variant,
  Phi     = summary_tbl$Phi,
  Phi_SE  = summary_tbl$Phi_SE
)

ggplot(phi_df, aes(x = reorder(Variant, Phi), y = Phi)) +
  geom_col(fill = "#1F77B4", width = 0.6) +
  geom_errorbar(aes(ymin = Phi - Phi_SE, ymax = Phi + Phi_SE),
                width = 0.25, colour = "grey30") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title    = "Nogueira size-corrected stability per variant",
       subtitle = "Error bars = ±1 SE  |  Dashed line = random-selection null (Φ̂ = 0)",
       x = NULL, y = expression(hat(Phi))) +
  theme_bw()
```

```{r show-jaccard, echo=FALSE}
#| fig-cap: "Mean Jaccard similarity of stable sets across the five validation folds. High values indicate SF selects the same genes regardless of which patients it trains on."
if (!is.null(sf_cache)) print(sf_cache$p_jaccard)
```

### Final models and held-out test evaluation

We refit **every** variant on the full train+validation set at its locked
hyperparameters, then score each on the vaulted test set — the single time
the test data is touched. Because variant selection was committed on the
validation folds, each test AUC is an honest estimate of that variant's
generalisation (`best_v` is simply the one we'd deploy). Each
`final_models[[v]]$fit` is a complete `mossyforest` object you can pass
straight to `plot_stability()`, `plot_importance()`, or mine for
`final_SHAP` / `feature_list` / `stability_freq`.

```{r final-models, eval=FALSE}
final_models <- lapply(names(VARIANTS), function(vname) {
  variant <- VARIANTS[[vname]]
  row     <- tune[[vname]]$best_row

  sc <- screen_control(
    drop_fraction = row$screen_drop,
    min_ntree     = 500L
  )
  sl <- select_control(
    drop_fraction    = row$select_drop,
    pi_thr           = row$pi_thr,
    n_boots          = 25L,
    min_ntree        = 500L,
    shadow_mode      = variant$shadow_mode,
    use_cfres = variant$use_cfres,
    cfres_scope        = variant$cfres_scope
  )

  set.seed(42)
  fit <- mf(X_tv, y_tv,
            module_membership = module_membership,   # global, from WGCNA section
            screen_params     = sc,
            select_params     = sl,
            backend           = "python",
            seed              = 42L,
            verbose           = 0L)

  stable <- fit$feature_list$feature_name
  rf <- ranger::ranger(
    y ~ ., probability = TRUE, num.trees = 300L, seed = 1L,
    data = data.frame(y = factor(y_tv), X_tv[, stable, drop = FALSE])
  )
  prob <- predict(rf, data.frame(X_test[, stable, drop = FALSE]))$predictions[, 2]

  list(
    fit        = fit,                       # full SF object for your own plots
    test_auc   = as.numeric(pROC::auc(y_test, prob, quiet = TRUE)),
    test_brier = mean((prob - y_test)^2),
    n_stable   = length(stable)
  )
})
names(final_models) <- names(VARIANTS)

# Honest test-vault performance per variant (selected variant flagged)
test_tbl <- data.frame(
  Variant    = names(final_models),
  Test_AUC   = round(vapply(final_models, `[[`, numeric(1), "test_auc"),   3),
  Test_Brier = round(vapply(final_models, `[[`, numeric(1), "test_brier"), 3),
  n_stable   = vapply(final_models, `[[`, integer(1), "n_stable"),
  Selected   = ifelse(names(final_models) == best_v, "<<", ""),
  row.names  = NULL
)
knitr::kable(test_tbl, align = "lcccc",
             caption = "Held-out test-vault performance (selection made on validation only).")
```

```{r show-test, echo=FALSE}
if (!is.null(sf_cache))
  knitr::kable(sf_cache$test_tbl, align = "lcccc",
               caption = "Held-out test-vault performance (selection made on validation only).")
```

The test vault is small (~15% of an already small cohort), so read these
numbers as noisy point estimates — the validation-fold distributions and the
pairwise Wilcoxon table above remain the stronger basis for *comparing*
variants. The vault's job is to hand the deployed model one selection-free
generalisation figure.

### Biological validation

For the BCR/ABL-vs-NEG contrast we use an **annotation-free** sanity check: do
the genes SF deems stable line up with the genes that simply separate the two
subtypes best on their own (largest |t-statistic|)? If the deployed stable set
recovers most of the top univariate hits, SF is capturing the dominant
molecular-subtype signal rather than fitting noise — and it does so while
controlling for the module-correlation structure a plain t-test ignores.

```{r bio, eval=FALSE}
cat("Selected variant:", best_v, "\n")

# Deployed stable set (selected variant, refit on full train+val)
stable_final <- final_models[[best_v]]$fit$feature_list$feature_name
cat("Deployed stable set size:", length(stable_final), "\n")

# Robustness view: union of the selected variant's stable sets across val folds
stable_union <- Reduce(union, tune[[best_v]]$sets)
cat("Union stable set across validation folds:", length(stable_union), "\n")

# Recovery of the strongest univariate BCR/ABL-vs-NEG signal (annotation-free):
# how many of the top-30 |t-stat| probes land in the deployed stable set?
tstats  <- apply(X_tv, 2, function(col)
  t.test(col[y_tv == 1], col[y_tv == 0])$statistic)
top_t   <- names(sort(abs(tstats), decreasing = TRUE))[1:30]
cat("Top-30 |t-stat| probes recovered in deployed set:",
    length(intersect(top_t, stable_final)), "/ 30\n")
```

```{r save-results, eval=FALSE}
library(ggplot2)
saveRDS(
  list(
    p_stability  = plot_stability(fit),
    p_importance = plot_importance(fit, top_n = 20),
    p_jaccard    = ggplot(phi_df, aes(x = reorder(Variant, Phi), y = Phi)) +
                     geom_col(fill = "#1F77B4", width = 0.6) +
                     geom_errorbar(aes(ymin = Phi - Phi_SE, ymax = Phi + Phi_SE),
                                   width = 0.25, colour = "grey30") +
                     geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
                     coord_flip() +
                     labs(title    = "Nogueira size-corrected stability per variant",
                          subtitle = "Error bars = ±1 SE  |  Dashed line = random-selection null (Φ̂ = 0)",
                          x = NULL, y = expression(hat(Phi))) +
                     theme_bw(),
    summary_tbl  = summary_tbl,
    pw           = pw,
    test_tbl     = test_tbl
  ),
  "leukemia_results.rds"
)
```

---

# Reporting Checklist

| Item | Source |
|------|--------|
| Validation-fold AUC + stability per variant | `summary_tbl` |
| Held-out test AUC / Brier per variant | `test_tbl` |
| Selected variant (1-SE rule) | `best_v` |
| Locked tuned hyperparameters | `summary_tbl` Best * columns |
| Pairwise variant comparison | paired Wilcoxon + BH table |
| Jaccard across validation folds | bar chart |
| Soft-threshold power justification | R² + connectivity plots |
| Biological overlap | deployed stable set vs known markers |
| Fitted models for plotting | `final_models[[v]]$fit` |

---

# Session Info

```{r session}
sessionInfo()
```
