---
title: "ShapleyForest on Simulated Data: Module-Based DGM"
author: "Timothy Shen"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: true
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Shapley Forest on Simulated Data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 9, fig.height = 5)
```

# Introduction

This vignette runs the full Shapley Forest (SF) pipeline on a **synthetic dataset
with known ground truth**, allowing direct evaluation of whether SF selects the
correct features rather than merely stable or predictive ones.

The vignette is structured in four stages:

1. Data generation and ground truth definition
2. WGCNA module detection
3. Single SF run and output for exploration purposes
4. Six-variant cross-validated comparison with ground-truth evaluation

---

# Data Generation

## Simulation design

We generate a co-expression network with **10 modules of 100 features each**
(1 000 features total) using a preferential-attachment graph per module,
mirroring the benchmark DGM used to evaluate SF variants.

**Network structure.** Modules 1–9 are correlated: within each module, a
Barabási–Albert preferential-attachment graph is built, its weighted adjacency
matrix forms the covariance, and features are sampled from
$\mathcal{MVN}(0, \Sigma)$.  Module 10 is fully independent ($\text{iid}\;\mathcal{N}(0,1)$).
Sparse random edges between modules add realistic cross-module bleed.

**Ground truth.** Three features from each of two correlated modules (M1, M2)
and three from the independent module (M10) are causal — nine signals total, all
carrying the same coefficient magnitudes $\boldsymbol{\beta} = (12, 12, 10)$.
The equal signal magnitude across correlated and independent sources directly
tests whether SF's split-pool and DML designs help it recover features from
both sources simultaneously.

```{r packages}
#| warning: false
#| message: false

if (!require("devtools", quietly = TRUE))
    install.packages("devtools")
library(devtools)
devtools::install_github("timothyzshen/shapleyforest")

library(shapleyforest)
library(WGCNA)
library(igraph)
library(MASS)
library(Matrix)
```

```{r dgm-fns}
# ── Network X generator: PA topology + eigengene loadings ────────────────────
# Each correlated module has a latent eigengene E_k ~ N(0,1).
# Feature j loads on E_k with strength proportional to its PA degree:
#   X_{k,j} = λ_j * E_k + sqrt(1 - λ_j²) * ε_j,  ε_j ~ N(0,1)
# giving Cor(X_{k,i}, X_{k,j}) = λ_i * λ_j — strong for hub pairs, weak for
# peripheral pairs, exactly the structure WGCNA's TOM is designed to detect.
# Causal features are sampled at random positions, independent of hub status.
generate_X <- function(
    n,
    n_modules   = 10L,   # total modules; last is always the independent pool
    module_size = 100L,  # features per module (equal across all modules)
    lambda_min  = 0.5,   # eigengene loading for lowest-degree (peripheral) nodes
    lambda_max  = 0.9,   # eigengene loading for highest-degree (hub) nodes
    seed        = NULL
) {
  if (!is.null(seed)) set.seed(seed)

  n_corr <- n_modules - 1L
  p_corr <- n_corr * module_size

  X_corr     <- matrix(0.0, n, p_corr)
  lambdas    <- numeric(p_corr)
  causal_pos <- vector("list", n_corr)

  for (k in seq_len(n_corr)) {
    m_k  <- min(3L, max(1L, floor(module_size / 10L)))
    gk   <- igraph::sample_pa(module_size, power = 1L, m = m_k, directed = FALSE)
    deg  <- igraph::degree(gk)

    # Loadings scale monotonically with degree: leaf → lambda_min, hub → lambda_max
    lambda_k <- lambda_min + (lambda_max - lambda_min) *
                (deg - min(deg)) / max(max(deg) - min(deg), 1L)

    E_k  <- rnorm(n)   # latent module eigengene
    cols <- ((k - 1L) * module_size + 1L):(k * module_size)

    eps  <- matrix(rnorm(n * module_size), n, module_size)
    eps  <- t(t(eps) * sqrt(pmax(1 - lambda_k^2, 1e-8)))  # per-feature noise SD
    X_corr[, cols] <- outer(E_k, lambda_k) + eps
    lambdas[cols]  <- lambda_k

    # Causal features: 3 random positions, independent of hub status
    causal_pos[[k]] <- sort(sample(module_size, 3L))
  }

  X_corr <- scale(X_corr)

  # Independent module: pure iid N(0, 1)
  X_indep      <- scale(matrix(rnorm(n * module_size), n, module_size))
  causal_indep <- sort(sample(module_size, 3L))

  # Module-aware column names: M1_V1 … M10_V100
  cn_corr  <- unlist(lapply(seq_len(n_corr),
    function(k) paste0("M", k, "_V", seq_len(module_size))))
  cn_indep <- paste0("M", n_modules, "_V", seq_len(module_size))
  colnames(X_corr)  <- cn_corr
  colnames(X_indep) <- cn_indep
  X <- as.data.frame(cbind(X_corr, X_indep))

  membership <- setNames(
    c(rep(paste0("M", seq_len(n_corr)), each = module_size),
      rep(paste0("M", n_modules, "_indep"), module_size)),
    colnames(X)
  )

  T_large <- cn_corr[(0L * module_size) + causal_pos[[1L]]]
  T_small <- cn_corr[(1L * module_size) + causal_pos[[2L]]]
  T_ind   <- cn_indep[causal_indep]

  list(
    X          = X,
    membership = membership,
    T_large    = T_large,
    T_small    = T_small,
    T_ind      = T_ind,
    T_all      = c(T_large, T_small, T_ind),
    lambdas    = setNames(c(lambdas, rep(NA_real_, module_size)), colnames(X))
  )
}

# ── DGM 1: pure linear, β = (12, 12, 10) per triplet ────────────────────────
make_y_linear <- function(X, T_large, T_small, T_ind, noise_sd = 1.0, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  Xm   <- as.matrix(X)
  beta <- setNames(rep(0, ncol(Xm)), colnames(Xm))
  beta[T_large] <- c(12, 12, 10)
  beta[T_small] <- c(12, 12, 10)
  beta[T_ind]   <- c(12, 12, 10)
  as.numeric(Xm %*% beta + rnorm(nrow(Xm), 0, noise_sd))
}
```

```{r dgm}
if (file.exists("simulation_data.rds")) {
  .d      <- readRDS("simulation_data.rds")
  sim_X        <- .d$sim_X
  sim_y        <- .d$sim_y
  true_modules <- .d$true_modules
  T_large      <- .d$T_large
  T_small      <- .d$T_small
  T_ind        <- .d$T_ind
  T_all        <- .d$T_all
  rm(.d)
} else {
  dgm <- generate_X(
    n           = 200L,     # ≥200 required for reliable WGCNA module detection
    n_modules   = 10L,
    module_size = 100L,
    seed        = 2026L
  )
  sim_X        <- dgm$X
  true_modules <- dgm$membership
  T_large      <- dgm$T_large    # M1 causal: 3 randomly sampled positions
  T_small      <- dgm$T_small    # M2 causal: 3 randomly sampled positions
  T_ind        <- dgm$T_ind      # M10 causal: 3 randomly sampled positions
  T_all        <- dgm$T_all      # all 9 signal features

  y_cont <- make_y_linear(sim_X, T_large, T_small, T_ind, noise_sd = 1.0, seed = 2026L)
  sim_y  <- as.integer(y_cont > median(y_cont))
}
```

```{r save-sim-data, eval=FALSE}
saveRDS(
  list(sim_X = sim_X, sim_y = sim_y, true_modules = true_modules,
       T_large = T_large, T_small = T_small, T_ind = T_ind, T_all = T_all),
  "simulation_data.rds"
)
```

```{r wgcna-run}
# Soft-thresholding power: lowest power where scale-free R² ≥ 0.80
sft <- WGCNA::pickSoftThreshold(
  sim_X,
  powerVector = c(1:10, seq(12, 20, 2)),
  verbose     = 0,
  networkType = "signed"
)
power_sel <- with(sft$fitIndices,
  Power[which(SFT.R.sq >= 0.80 & slope < 0)[1]])
if (is.na(power_sel)) power_sel <- 12L
cat("Soft-thresholding power:", power_sel, "\n")

net <- WGCNA::blockwiseModules(
  sim_X,
  power          = power_sel,
  networkType    = "signed",
  TOMType        = "signed",
  minModuleSize  = 30L,
  mergeCutHeight = 0.25,
  numericLabels  = FALSE,
  verbose        = 0
)
wgcna_colors <- net$colors
cat("WGCNA modules :", length(unique(wgcna_colors)) - 1L, "(excl. grey)\n")
cat("Grey features  :", sum(wgcna_colors == "grey"), "/", ncol(sim_X), "\n")
```

```{r wgcna-compare}
# For each true correlated module, find best-matching WGCNA colour and
# compute precision/recall. M10_indep features are expected to go grey.
recovery_tbl <- do.call(rbind, lapply(
  setdiff(unique(true_modules), "M10_indep"),
  function(tm) {
    true_feat <- names(true_modules)[true_modules == tm]
    best_col  <- names(which.max(table(wgcna_colors[true_feat])))
    recovered <- names(wgcna_colors)[wgcna_colors == best_col]
    tp        <- length(intersect(true_feat, recovered))
    data.frame(
      True_module = tm,
      WGCNA_color = best_col,
      True_n      = length(true_feat),
      WGCNA_n     = length(recovered),
      Overlap     = tp,
      Precision   = round(tp / length(recovered), 3),
      Recall      = round(tp / length(true_feat),  3),
      row.names   = NULL
    )
  }
))
knitr::kable(recovery_tbl, align = "lcccccc",
             caption = "WGCNA module recovery vs. true modules (M10_indep expected grey).")
```

```{r module-membership-true}
# Use the ground-truth module labels directly — the same approach as the
# benchmark (pilot_v3.R passes dat$membership straight to sf()).
# WGCNA is shown below for exploration only; with n=200 and a PA network it
# cannot reliably recover the true blocks.
module_membership <- true_modules
# SF identifies the independent pool by name:
module_membership[module_membership == "M10_indep"] <- "M10_indep"  # kept as-is
```

```{r sim-summary}
cat("Samples :", nrow(sim_X), "\n")
cat("Features:", ncol(sim_X), "  (9 correlated modules + 1 independent × 100 each)\n")
cat("Causal  :", length(T_all), "features —",
    paste(T_all, collapse = ", "), "\n")
cat("Outcome :", paste(names(table(sim_y)), collapse = " / "),
    "→", paste(table(sim_y), collapse = " / "), "\n")
```

```{r results-cache, echo=FALSE}
sim_cache <- if (file.exists("simulation_results.rds")) readRDS("simulation_results.rds") else NULL
```

The nine causal features carry equal signal magnitude whether they sit inside a
correlated module (M1, M2) or the fully independent module (M10).  A method
that handles only one pool will miss the other.

---

# Single SF Run

We fit SF once on the full dataset to explore its output before the formal
comparison.

```{r sf-setup, eval=FALSE}
sf_setup()
```

```{r single-run, eval=FALSE}
sc_single <- screen_control(min_ntree = 500L)
sl_single <- select_control(min_ntree = 500L, n_boots = 50L,
                             indep_modules = "M10_indep")

fit <- sf(
  X                 = sim_X,
  y                 = sim_y,
  module_membership = module_membership,
  screen_params     = sc_single,
  select_params     = sl_single,
  backend           = "python",
  verbose           = 1L,
  seed              = 2026L
)
```

## Stable set

```{r stable-set, eval=FALSE}
cat("Stable features selected:", length(fit$feature_list$feature_name), "\n")
head(sort(fit$stability_freq, decreasing = TRUE), 20)
```

### Ground-truth validation

Because the causal features are known, we evaluate SF's stable set directly
against the nine true signals.  We report overall recall and precision, plus
per-source recall to reveal whether SF recovers correlated-module features,
independent-module features, or both.

```{r gt-single, eval=FALSE}
stable <- fit$feature_list$feature_name

tp <- length(intersect(stable, T_all))
cat(sprintf("Overall\n"))
cat(sprintf("  M1 (corr large): %d / 3\n", length(intersect(stable, T_large))))
cat(sprintf("  M2 (corr small): %d / 3\n", length(intersect(stable, T_small))))
cat(sprintf("  M10 (indep)    : %d / 3\n", length(intersect(stable, T_ind))))
```

## Stability plot

```{r plot-stability, eval=FALSE}
#| fig-cap: "Shadow-win frequency per feature. Filled points exceed the stability threshold and enter the stable set."
plot_stability(fit)
```

```{r show-stability, echo=FALSE}
#| fig-cap: "Shadow-win frequency per feature. Filled points exceed the stability threshold and enter the stable set."
if (!is.null(sim_cache)) print(sim_cache$p_stability)
```


---

# Model Comparison: Three-Way Split + Cross-Validated Tuning

## Design

We compare **six SF variants** formed by two shadow modes × three DML options:

| Variant | Shadow mode | DML |
|---------|-------------|-----|
| `split` | split | none |
| `split_dml` | split | standard |
| `split_mdml` | split | marginal |
| `within` | within_module | none |
| `within_dml` | within_module | standard |
| `within_mdml` | within_module | marginal |

We proceed with the following:

0. **Modules from all data.** The `module_membership` from the section above is
   reused throughout.
1. **Vault a test set** (~20%, stratified) — held out once, scored once, at
   the very end.
2. **On the remaining ~80% (train):** 5-fold CV tunes the hyperparameter grid
   *and* compares the six variants on both performance (mean validation-fold AUC)
   **and** stability (Jaccard of the stable set across folds).
3. **Select one variant** by a pre-specified rule (below), refit every variant
   on the *full* train set, and evaluate on both the test vault and the
   ground-truth causal features.

Hyperparameters tuned jointly on the validation folds:

| Parameter | Grid | Default |
|-----------|------|---------|
| `screen_control(drop_fraction)` | 0.25 (fixed) | 0.25 |
| `select_control(drop_fraction)` | 0.05 (fixed) | 0.10 |
| `select_control(pi_thr)` | 0.50, 0.60, 0.70 | 0.60 |
| `min_ntree` (screen + select) | 500 (fixed) | 500 |
| `n_boots` | 25 (fixed) | 50 |

9 combinations per variant; selection criterion **mean validation-fold AUC**.
All other parameters stay at package defaults.

**Selection rule (fixed in advance).** Among the six variants we apply the
**one-standard-error rule**: take the best mean CV AUC, then among all
variants within one standard error of it, choose the **most stable** (highest
mean pairwise Jaccard).

## Setup

```{r cv-load-pack}
#| warning: false
#| message: false

library(pROC)
library(caret)
```

```{r cv-setup, eval=FALSE}
VARIANTS <- list(
  split       = list(shadow_mode = "split",         use_dml_residual = FALSE, dml_scope = "indep"),
  split_dml   = list(shadow_mode = "split",         use_dml_residual = TRUE,  dml_scope = "indep"),
  split_mdml  = list(shadow_mode = "split",         use_dml_residual = TRUE,  dml_scope = "all_modules"),
  within      = list(shadow_mode = "within_module", use_dml_residual = FALSE, dml_scope = "indep"),
  within_dml  = list(shadow_mode = "within_module", use_dml_residual = TRUE,  dml_scope = "indep"),
  within_mdml = list(shadow_mode = "within_module", use_dml_residual = TRUE,  dml_scope = "all_modules")
)

GRID <- expand.grid(
  screen_drop = 0.25,
  select_drop = 0.05,
  pi_thr      = c(0.50, 0.60, 0.70),
  stringsAsFactors = FALSE
)

set.seed(2026)
trainval_idx <- as.integer(caret::createDataPartition(
  factor(sim_y), p = 0.8, list = FALSE))
test_idx     <- setdiff(seq_along(sim_y), trainval_idx)

X_tv   <- sim_X[trainval_idx, ]; y_tv   <- sim_y[trainval_idx]
X_test <- sim_X[test_idx, ];     y_test <- sim_y[test_idx]

val_folds <- caret::createFolds(y_tv, k = 5, returnTrain = FALSE)

cat(sprintf("Train: %d  |  Test vault: %d  |  Test positives: %d\n",
            length(y_tv), length(y_test), sum(y_test == 1)))
```

## Fit-and-evaluate helper

```{r cv-helper, eval=FALSE}
fit_eval <- function(X_tr, y_tr, X_te, y_te, variant, grid_row) {

  sc  <- screen_control(
    drop_fraction = grid_row$screen_drop,
    min_ntree     = 500L
  )
  sl  <- select_control(
    drop_fraction    = grid_row$select_drop,
    pi_thr           = grid_row$pi_thr,
    n_boots          = 50L,
    min_ntree        = 500L,
    indep_modules    = "M10_indep",
    shadow_mode      = variant$shadow_mode,
    use_dml_residual = variant$use_dml_residual,
    dml_scope        = variant$dml_scope
  )

  fit <- tryCatch(
    sf(X_tr, y_tr,
       module_membership = module_membership,
       screen_params     = sc,
       select_params     = sl,
       backend           = "python",
       seed              = 2026L,
       verbose           = 0L),
    error = function(e) NULL
  )

  stable <- if (!is.null(fit)) fit$feature_list$feature_name else character(0)
  if (length(stable) == 0L)
    return(list(auc = NA_real_, brier = NA_real_,
                n_stable = 0L, stable = character(0)))

  rf   <- ranger::ranger(
    y ~ ., probability = TRUE, num.trees = 300L, seed = 1L,
    data = data.frame(y = factor(y_tr), X_tr[, stable, drop = FALSE])
  )
  prob <- predict(rf,
    data.frame(X_te[, stable, drop = FALSE]))$predictions[, 2]

  list(
    auc      = as.numeric(pROC::auc(y_te, prob, quiet = TRUE)),
    brier    = mean((prob - y_te)^2),
    n_stable = length(stable),
    stable   = stable
  )
}
```

## Smoke test

The full tuning sweep is 6 variants × 3 grid rows × 5 validation folds.
Before launching, verify that `fit_eval()` runs end-to-end on two grid rows
and a single train/test split.

```{r cv-smoke-test, eval=FALSE}
GRID_smoke <- GRID[1:2, ]
print(GRID_smoke)

set.seed(1)
smoke_idx <- caret::createFolds(sim_y, k = 5, returnTrain = FALSE)[[1]]
s_tr <- setdiff(seq_along(sim_y), smoke_idx)
s_te <- smoke_idx

smoke_results <- lapply(seq_len(nrow(GRID_smoke)), function(g) {
  fit_eval(sim_X[s_tr, ], sim_y[s_tr],
           sim_X[s_te, ], sim_y[s_te],
           VARIANTS[["split"]], GRID_smoke[g, ])
})

data.frame(
  grid_row = seq_len(nrow(GRID_smoke)),
  auc      = vapply(smoke_results, `[[`, numeric(1), "auc"),
  brier    = vapply(smoke_results, `[[`, numeric(1), "brier"),
  n_stable = vapply(smoke_results, `[[`, numeric(1), "n_stable")
)
```

If both rows return a finite `auc` and `n_stable > 0`, `fit_eval()` and
`sf()` are wired correctly and it is safe to launch the full `tune` sweep below.

## Cross-validated tuning and stability

For each variant we sweep all 9 grid rows across the five validation folds,
record each row's mean validation AUC, and lock the best-scoring row. At that
locked setting we also keep the per-fold stable sets, whose mean pairwise
Jaccard is the variant's **stability**. Nothing here touches the test vault.

```{r cv-tune, eval=FALSE}
jac_fn <- function(a, b)
  length(intersect(a, b)) / max(length(union(a, b)), 1L)

tune_one <- function(vname) {
  variant <- VARIANTS[[vname]]

  grid_res <- lapply(seq_len(nrow(GRID)), function(g) {
    per_fold <- lapply(seq_along(val_folds), function(fi) {
      vi   <- val_folds[[fi]]
      i_te <- vi; i_tr <- setdiff(seq_along(y_tv), vi)
      r <- fit_eval(X_tv[i_tr, ], y_tv[i_tr],
                    X_tv[i_te, ], y_tv[i_te],
                    variant, GRID[g, ])
      message(sprintf("[%s] g%02d fold%d — auc=%.3f n=%d",
                      vname, g, fi, r$auc, r$n_stable))
      r
    })
    list(
      auc  = mean(vapply(per_fold, `[[`, numeric(1), "auc"), na.rm = TRUE),
      aucs = vapply(per_fold, `[[`, numeric(1), "auc"),
      sets = lapply(per_fold, `[[`, "stable")
    )
  })

  best_g <- which.max(vapply(grid_res, `[[`, numeric(1), "auc"))
  best   <- grid_res[[best_g]]

  jac <- mean(combn(length(best$sets), 2,
                    function(i) jac_fn(best$sets[[i[1]]], best$sets[[i[2]]]),
                    simplify = TRUE))

  message(sprintf(">>> [%s] DONE — best g%02d cv_auc=%.3f jaccard=%.3f",
                  vname, best_g, best$auc, jac))

  list(
    variant  = vname,
    best_row = GRID[best_g, ],
    cv_auc   = best$auc,
    cv_aucs  = best$aucs,
    jaccard  = jac,
    sets     = best$sets
  )
}

tune <- lapply(names(VARIANTS), tune_one)
names(tune) <- names(VARIANTS)
```

## Results

### Variant selection

For each variant, hyperparameters were tuned in the CV sweep above by
predictive performance (CV AUC).  Variant selection itself uses a different
criterion: the **Nogueira size-corrected stability index** $\hat{\Phi}$
(Nogueira, Sechidis & Brown, 2018), which corrects for the inflation that
raw Jaccard similarity exhibits when comparing variants with different stable
set sizes.  Among variants whose $\hat{\Phi}$ falls within one SE of the
maximum, we select the most **parsimonious** (smallest mean stable set size
$\bar{k}$).  Predictive performance is reported for reference but does not
drive variant selection.

```{r cv-select, eval=FALSE}
# ── Nogueira Φ̂ helpers ───────────────────────────────────────────────────────
.phi_hat <- function(sets, d) {
  M     <- length(sets)
  k_bar <- mean(sapply(sets, length))
  if (k_bar == 0 || k_bar >= d) return(NA_real_)
  pairs <- combn(M, 2, simplify = FALSE)
  sym_d <- sapply(pairs, function(idx) {
    a <- sets[[idx[1]]]; b <- sets[[idx[2]]]
    length(a) + length(b) - 2L * length(intersect(a, b))
  })
  v_obs  <- mean(sym_d) / d
  v_null <- 2 * k_bar * (d - k_bar) / d^2
  1 - v_obs / v_null
}

.phi_se <- function(sets, d) {
  M     <- length(sets)
  k_bar <- mean(sapply(sets, length))
  if (k_bar == 0 || k_bar >= d) return(NA_real_)
  v_null <- 2 * k_bar * (d - k_bar) / d^2
  pairs  <- combn(M, 2, simplify = FALSE)
  v_ij   <- sapply(pairs, function(idx) {
    a <- sets[[idx[1]]]; b <- sets[[idx[2]]]
    (length(a) + length(b) - 2L * length(intersect(a, b))) / d
  })
  sd(v_ij) / (sqrt(length(v_ij)) * v_null)
}

d_total <- ncol(sim_X)

summary_tbl <- do.call(rbind, lapply(tune, function(t) {
  phi    <- .phi_hat(t$sets, d_total)
  phi_se <- .phi_se(t$sets, d_total)
  k_bar  <- mean(sapply(t$sets, length))
  data.frame(
    Variant        = t$variant,
    CV_AUC         = round(t$cv_auc, 3),
    CV_AUC_SE      = round(sd(t$cv_aucs, na.rm = TRUE) /
                             sqrt(sum(!is.na(t$cv_aucs))), 3),
    Jaccard        = round(t$jaccard, 3),
    Phi            = round(phi, 3),
    Phi_SE         = round(phi_se, 3),
    k_bar          = round(k_bar, 1),
    `Best pi_thr`  = t$best_row$pi_thr,
    `Best sc_drop` = t$best_row$screen_drop,
    `Best sl_drop` = t$best_row$select_drop,
    check.names = FALSE, row.names = NULL
  )
}))

# 1-SE rule on Φ̂; parsimony tiebreaker (smallest k̄)
top     <- which.max(summary_tbl$Phi)
cutoff  <- summary_tbl$Phi[top] - summary_tbl$Phi_SE[top]
in_band <- summary_tbl[summary_tbl$Phi >= cutoff, ]
best_v  <- in_band$Variant[which.min(in_band$k_bar)]

knitr::kable(summary_tbl, align = "lccccccccc",
             caption = "Per-variant CV performance and Nogueira stability (Φ̂). Variant selection uses 1-SE rule on Φ̂ with parsimony tiebreaker.")
cat("Selected variant (1-SE on Φ̂, parsimony tiebreak):", best_v, "\n")
```

```{r show-summary, echo=FALSE}
if (!is.null(sim_cache))
  knitr::kable(sim_cache$summary_tbl, align = "lccccccccc",
               caption = "Per-variant CV performance and Nogueira stability (Φ̂). Variant selection uses 1-SE rule on Φ̂ with parsimony tiebreaker.")
```

### Pairwise variant comparison

The five validation folds give five paired AUC values per variant.  We test
all 15 pairs with a paired Wilcoxon test and apply Benjamini-Hochberg
correction.

```{r pairwise, eval=FALSE}
pairs <- combn(names(VARIANTS), 2, simplify = FALSE)

pw <- do.call(rbind, lapply(pairs, function(p) {
  a <- tune[[p[1]]]$cv_aucs
  b <- tune[[p[2]]]$cv_aucs
  wt <- wilcox.test(a, b, paired = TRUE, exact = FALSE)
  data.frame(`Method A` = p[1], `Method B` = p[2],
             `ΔAUC`     = round(mean(a - b, na.rm = TRUE), 3),
             `p-value`  = round(wt$p.value, 4),
             check.names = FALSE)
}))
pw$`p-adj (BH)` <- round(p.adjust(pw$`p-value`, method = "BH"), 4)
knitr::kable(pw, caption = "Pairwise Wilcoxon tests on validation-fold AUC.")
```

```{r show-pairwise, echo=FALSE}
if (!is.null(sim_cache))
  knitr::kable(sim_cache$pw, caption = "Pairwise Wilcoxon tests on validation-fold AUC.")
```

### Nogueira stability

```{r phi-plot, eval=FALSE}
#| fig-cap: "Nogueira size-corrected stability index (Φ̂) per variant. Error bars show ±1 SE. Dashed line = 0 (random-selection null). Unlike raw Jaccard, Φ̂ accounts for the expected overlap at each variant's selection size, enabling unbiased cross-variant comparison."
library(ggplot2)

phi_df <- data.frame(
  Variant = summary_tbl$Variant,
  Phi     = summary_tbl$Phi,
  Phi_SE  = summary_tbl$Phi_SE
)

ggplot(phi_df, aes(x = reorder(Variant, Phi), y = Phi)) +
  geom_col(fill = "#1F77B4", width = 0.6) +
  geom_errorbar(aes(ymin = Phi - Phi_SE, ymax = Phi + Phi_SE),
                width = 0.25, colour = "grey30") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title    = "Nogueira size-corrected stability per variant",
       subtitle = "Error bars = ±1 SE  |  Dashed line = random-selection null (Φ̂ = 0)",
       x = NULL, y = expression(hat(Phi))) +
  theme_bw()
```

```{r show-jaccard, echo=FALSE}
#| fig-cap: "Nogueira size-corrected stability index (Φ̂) per variant with ±1 SE error bars."
if (!is.null(sim_cache)) print(sim_cache$p_jaccard)
```

### Final models and held-out test evaluation

We refit **every** variant on the full train+validation set at its locked
hyperparameters, then score each on the vaulted test set — the single time
the test data is touched.  Each `final_models[[v]]$fit` is a complete
`shapleyforest` object available for further exploration.

```{r final-models, eval=FALSE}
final_models <- lapply(names(VARIANTS), function(vname) {
  variant <- VARIANTS[[vname]]
  row     <- tune[[vname]]$best_row

  sc <- screen_control(
    drop_fraction = row$screen_drop,
    min_ntree     = 500L
  )
  sl <- select_control(
    drop_fraction    = row$select_drop,
    pi_thr           = row$pi_thr,
    n_boots          = 25L,
    min_ntree        = 1000L,
    indep_modules    = "M10_indep",
    shadow_mode      = variant$shadow_mode,
    use_dml_residual = variant$use_dml_residual,
    dml_scope        = variant$dml_scope
  )

  fit <- sf(X_tv, y_tv,
            module_membership = module_membership,
            screen_params     = sc,
            select_params     = sl,
            backend           = "python",
            seed              = 2026L,
            verbose           = 0L)

  stable <- fit$feature_list$feature_name
  rf <- ranger::ranger(
    y ~ ., probability = TRUE, num.trees = 300L, seed = 1L,
    data = data.frame(y = factor(y_tv), X_tv[, stable, drop = FALSE])
  )
  prob <- predict(rf, data.frame(X_test[, stable, drop = FALSE]))$predictions[, 2]

  list(
    fit        = fit,
    test_auc   = as.numeric(pROC::auc(y_test, prob, quiet = TRUE)),
    test_brier = mean((prob - y_test)^2),
    n_stable   = length(stable)
  )
})
names(final_models) <- names(VARIANTS)

test_tbl <- data.frame(
  Variant    = names(final_models),
  Test_AUC   = round(vapply(final_models, `[[`, numeric(1), "test_auc"),   3),
  Test_Brier = round(vapply(final_models, `[[`, numeric(1), "test_brier"), 3),
  n_stable   = vapply(final_models, `[[`, integer(1), "n_stable"),
  Selected   = ifelse(names(final_models) == best_v, "<<", ""),
  row.names  = NULL
)
knitr::kable(test_tbl, align = "lcccc",
             caption = "Held-out test-vault performance (selection made on validation only).")
```

```{r show-test, echo=FALSE}
if (!is.null(sim_cache))
  knitr::kable(sim_cache$test_tbl, align = "lcccc",
               caption = "Held-out test-vault performance (selection made on validation only).")
```

### Ground-truth feature recovery

The key evaluation unique to this simulation: for each variant's final model,
how many of the nine causal features are recovered?  Recall and precision are
computed against the known `T_all` set.  The correlated-module columns (M1, M2)
and independent-module column (M10) are reported separately to reveal whether
SF's split-pool and DML designs help at recovering each source.

```{r gt-final, eval=FALSE}
cat("Selected variant:", best_v, "\n\n")

gt_tbl <- do.call(rbind, lapply(names(final_models), function(v) {
  s  <- final_models[[v]]$fit$feature_list$feature_name
  tp <- length(intersect(s, T_all))
  data.frame(
    Variant    = v,
    Recall_M1  = length(intersect(s, T_large)),
    Recall_M2  = length(intersect(s, T_small)),
    Recall_M10 = length(intersect(s, T_ind)),
    Total_TP   = tp,
    Recall     = round(tp / length(T_all), 3),
    Precision  = round(tp / max(length(s), 1L), 3),
    n_stable   = length(s),
    Selected   = ifelse(v == best_v, "<<", ""),
    row.names  = NULL
  )
}))
knitr::kable(gt_tbl, align = "lcccccccc",
             caption = "Ground-truth feature recovery per variant (final models, full train set).")
```

```{r show-gt, echo=FALSE}
if (!is.null(sim_cache))
  knitr::kable(sim_cache$gt_tbl, align = "lcccccccc",
               caption = "Ground-truth feature recovery per variant (final models, full train set).")
```

### Bootstrap selection frequency

The final model is a single fit on `X_tv`.  To assess which features are
*consistently* selected across different subsets of patients — not just in one
run — we repeat the selected variant at its locked hyperparameters on
`B = 100` random 80% subsamples of the training set (without replacement).
The resulting selection frequency for each feature is the fraction of runs in
which it entered the stable set.  Because we aggregate across 100 runs, each
individual fit uses `n_boots = 15` (fewer bootstraps per run) without
materially affecting the aggregate frequency estimate.

```{r boot-stability, eval=FALSE}
row_best     <- tune[[best_v]]$best_row
variant_best <- VARIANTS[[best_v]]

best_sc <- screen_control(
  drop_fraction = row_best$screen_drop,
  min_ntree     = 500L
)
best_sl <- select_control(
  drop_fraction    = row_best$select_drop,
  pi_thr           = row_best$pi_thr,
  n_boots          = 15L,
  min_ntree        = 500L,
  indep_modules    = "M10_indep",
  shadow_mode      = variant_best$shadow_mode,
  use_dml_residual = variant_best$use_dml_residual,
  dml_scope        = variant_best$dml_scope
)

B        <- 100L
sub_size <- floor(0.8 * nrow(X_tv))

boot_sets <- lapply(seq_len(B), function(b) {
  idx   <- sample(nrow(X_tv), sub_size, replace = FALSE)
  fit_b <- tryCatch(
    sf(X_tv[idx, ], y_tv[idx],
       module_membership = module_membership,
       screen_params     = best_sc,
       select_params     = best_sl,
       backend           = "python",
       seed              = b,
       verbose           = 0L),
    error = function(e) NULL
  )
  message(sprintf("boot %03d/%d — %s", b, B,
                  if (is.null(fit_b)) "FAILED"
                  else sprintf("n_stable=%d", nrow(fit_b$feature_list))))
  if (is.null(fit_b)) character(0) else fit_b$feature_list$feature_name
})

# Selection frequency per feature
freq_vec <- sort(
  sapply(colnames(X_tv), function(f) mean(sapply(boot_sets, function(s) f %in% s))),
  decreasing = TRUE
)
freq_df <- data.frame(
  feature   = names(freq_vec),
  frequency = as.numeric(freq_vec),
  source    = ifelse(names(freq_vec) %in% T_large, "M1 (causal)",
              ifelse(names(freq_vec) %in% T_small, "M2 (causal)",
              ifelse(names(freq_vec) %in% T_ind,   "M10 (causal)",
                                                    "Non-causal")))
)
freq_df <- freq_df[freq_df$frequency > 0, ]
```

```{r boot-freq-plot, eval=FALSE}
#| fig-cap: "Bootstrap selection frequency — top 10 features per module. Causal features are coloured by module; co-members selected incidentally are grey."
library(ggplot2)

# Identify which module label corresponds to each causal group
m1_mod  <- unique(module_membership[T_large])
m2_mod  <- unique(module_membership[T_small])
m10_mod <- unique(module_membership[T_ind])

feat_mod <- module_membership[freq_df$feature]
freq_df$panel <- ifelse(feat_mod == m1_mod,  "M1 module",
                 ifelse(feat_mod == m2_mod,  "M2 module",
                 ifelse(feat_mod == m10_mod, "M10 module", "Other modules")))
freq_df$panel <- factor(freq_df$panel,
  levels = c("M1 module", "M2 module", "M10 module", "Other modules"))

top_per_module <- do.call(rbind, lapply(levels(freq_df$panel), function(p) {
  sub <- freq_df[freq_df$panel == p, ]
  head(sub[order(-sub$frequency), ], 10)
}))
top_per_module$feature <- reorder(top_per_module$feature, top_per_module$frequency)

p_boot <- ggplot(top_per_module, aes(x = feature, y = frequency, fill = source)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("M1 (causal)"  = "#D62728",
                                "M2 (causal)"  = "#FF7F0E",
                                "M10 (causal)" = "#2CA02C",
                                "Non-causal"   = "#AEC7E8")) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ panel, scales = "free_y", ncol = 2) +
  coord_flip() +
  labs(title    = sprintf("Bootstrap selection frequency — %s", best_v),
       subtitle = sprintf("B = %d runs, 80%% subsample (n = %d)", B, sub_size),
       x = NULL, y = "Selection frequency", fill = "Feature type") +
  theme_bw()
print(p_boot)
```

```{r show-boot, echo=FALSE}
#| fig-cap: "Bootstrap selection frequency for the top 30 features of the selected variant."
if (!is.null(sim_cache)) print(sim_cache$p_boot)
```

```{r save-results, eval=FALSE}
library(ggplot2)
saveRDS(
  list(
    p_stability  = plot_stability(fit),
    p_importance = plot_importance(fit, top_n = 20),
    p_jaccard    = ggplot(phi_df, aes(x = reorder(Variant, Phi), y = Phi)) +
                     geom_col(fill = "#1F77B4", width = 0.6) +
                     geom_errorbar(aes(ymin = Phi - Phi_SE, ymax = Phi + Phi_SE),
                                   width = 0.25, colour = "grey30") +
                     geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
                     coord_flip() +
                     labs(title    = "Nogueira size-corrected stability per variant",
                          subtitle = "Error bars = ±1 SE  |  Dashed line = random-selection null (Φ̂ = 0)",
                          x = NULL, y = expression(hat(Phi))) +
                     theme_bw(),
    summary_tbl  = summary_tbl,
    pw           = pw,
    test_tbl     = test_tbl,
    gt_tbl       = gt_tbl,
    p_boot       = p_boot,
    freq_df      = freq_df
  ),
  "simulation_results.rds"
)
```

---

# Reporting Checklist

| Item | Source |
|------|--------|
| Validation-fold AUC + stability per variant | `summary_tbl` |
| Held-out test AUC / Brier per variant | `test_tbl` |
| Selected variant (1-SE rule) | `best_v` |
| Locked tuned hyperparameters | `summary_tbl` Best * columns |
| Pairwise variant comparison | paired Wilcoxon + BH table |
| Jaccard across validation folds | bar chart |
| Ground-truth recall per variant | `gt_tbl` |
| WGCNA module purity vs true modules | `module-recovery` chunk |
| Fitted models for plotting | `final_models[[v]]$fit` |

---

# Session Info

```{r session}
sessionInfo()
```
